\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-02-13 eg6-letters/NIM/EG6_RTPC_NIM.tex
abstract/introduction updates

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at Jefferson 
laboratory (JLab) to track low-energy nuclear recoils for the purpose of 
measuring nuclear exclusive channels, such as, coherent Deeply Virtual Compton 
Scattering (DVCS) and coherent production of mesons on $^4$He. In such 
processes, $^4$He nucleus remains intact and recoils as a whole in the final 
state. The CEBAF Large Acceptance Spectrometer (CLAS) at JLab has been used for 
similar studies on proton targets, however it cannot track recoils with momenta 
less than 250 MeV/c. In 2009, we carried out measurements on $^4$He target 
using CLAS spectrometer, which was supplemented with the RTPC during the 
CLAS-EG6 experimental run. The detector was positioned directly around a 
gaseous target, allowing a detection threshold as low as $~$100  MeV/c for 
$^4$He. This article discuses the RTPC design, work principle, calibration 
methods and performances.
\end{abstract}

@@

The CEBAF Large Acceptance Spectrometer (CLAS)~\cite{CLASref} is installed in 
the Hall-B of the Thomas Jefferson National Accelerator Facility (TJNAF), 
Newport News, Virginia, USA. The accelerator of TJNAF provides, simultaneously, 
a high power electron beam of energy up to 6 GeV and 100$\%$ duty cycle to 
three experimental Halls (A, B, C). The CLAS spectrometer, located in Hall-B, 
is composed of several sub-detectors and two magnets. The spectrometer was 
designed to track charged particles of momentum greater than 250 MeV/c as well 
as detecting real photons. Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item The three regions of Drift Chambers (DC) used for the tracking of charged 
    particles~\cite{DCref} (in blue on the figure).
 \item The superconducting toroidal magnet (in yellow) bends the trajectories 
    of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item The Cherenvkov Counters (CC) used to separate electrons from other negative 
    particles, such as pions, at lower energies~\cite{CCref} (in pink).
 \item The Scintillation Counters (SC) used to identify hadrons by measuring their 
    Time-Of-Flight (TOF)~\cite{TOFref} (in red).
 \item The Electromagnetic Calorimeter (EC) used to separate electrons from other negative 
    particles at higher energies and to detect neutral particles (neutrons and 
    photons in particular)~\cite{ECref} (in green).
\end{itemize}

@@

The measurement of the DVCS process ($eH \rightarrow e' H' \gamma$, where 
$H$ is a hadron) necessitates some adaptations in this setup. Indeed, the EC 
covers only polar angles from 8$^{\circ}$ to 45$^{\circ}$ with a very small
acceptance at the lowest angles. With a 6 GeV electron beam on a proton target, the 
majority of the DVCS photons are produced at polar angles below 15$^{\circ}$, 
where the acceptance of the EC is poor. To resolve this issue an additional 
Inner Calorimeter (IC) has been built for the E01-113 experiment (2005). The IC is 
constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$. Each crystal is 16 cm long,
corresponding to 17 radiation lengths. The achieved energy 
resolution is around 3 to 4$\%$ for photon energies between 2 GeV and 5 GeV and 
the angular resolution is between 3 to 5 mrad for the same energy range 
\cite{Hyon-suk}. The amount of background from Moller electron is however
very problematic at such small angles, so a 5 Tesla solenoid magnet has been
added around the target as a magnetic shield against the low energy background.

The developed RTPC has been used during a three months experimental run in 2009 
with a longitudinally polarized electron beam of energy 6.064 GeV projected 
onto a gaseous $^{4}$He target to measure asymmetries in Deep Virtual Compton 
Scattering (DVCS)~\cite{proposal} and exclusive meson production. Moreover, a 
5~T solenoid was placed around the RTPC-target system to shield the detectors 
close to the beam line from Mollers electrons. The solenoid field plays also a 
major rule in finding the momentum of the charged particles in the RTPC by 
bending their tracks curvature.

In this paper we report the progress in designing and calibrating a radial TPC 
that had been used with JLab-CLAS spectrometer. This paper is organized as 
follows: section \ref{prev_tpc} presents an overview of the TPCs history and 
previous usage, section \ref{sec_design} details the physics considerations for 
the RTPC design and its internal structure, while section \ref{sec_readout} 
points the properties of the read-out system. The calibration strategies for 
such a detector are discussed in section \ref{sec_calib}. Finally, the 
performance of the RTPC is described in section \ref{sec_perf}.

\section{TPC technology} \label{prev_tpc}
In the late 1970s, David R. Nygren presented the concept of the Time projection 
Chambers (TPCs) as a new tool to track charged particles based on combining 
electric and magnetic fields within a sensitive gas volume. The ionized 
electrons are collected by a position-sensitive electron collection system.  
Later in 1997, F.~Sauli introduced the Gas Electron Multiplier (GEM) as an 
effective gas-gain element to improve the tracking efficiency of the TPCs 
\ref{gem_sauli}. Nowadays, the TPCs are widely used in physics experiments, 
such as high-energy physics (ALEPH, DELPHI), heavy-ion physics (STAR, ALICE) 
and rare-event searches (XENON, NEXT).

The first implementation of a radial TPC with JLab-CLAS spectrometer has been 
performed by the BoNuS experiment \cite{BONUS} in 2005. The success of the 
BoNuS-RTPC in detecting the slow protons motivated the development of this new 
RTPC. For the channels of interest in CLAS-EG6 experiment, the recoiled 
$^{4}$He nuclei have momentum per charge around 150 MeV/c. The detection of 
these low-energy nuclei was the primary motivation for designing this detector.

>>>>

\begin{abstract}
A new Radial Time Projection Chamber (RTPC) was developed at Jefferson 
Laboratory (JLab) to track low-energy nuclear recoils for the purpose of 
measuring exclusive nuclear channels, such as coherent Deeply Virtual Compton 
Scattering (DVCS) and coherent meson production on $^4$He. In such 
processes, the $^4$He nucleus remains intact and recoils as a whole in the final 
state. The CEBAF Large Acceptance Spectrometer (CLAS) at JLab has been used for 
similar studies on proton targets, however it cannot track recoils with momenta 
less than 250 MeV/c. In 2009, we carried out measurements using the CLAS spectrometer supplemented by the RTPC positioned directly around a gaseous $^4$He target, allowing a detection threshold as low as $~$100  MeV/c for 
$^4$He. This article discuses the design, work principle, calibration 
methods and performances of the RTPC.
\end{abstract}

@@

Thomas Jefferson National Accelerator Facility (TJNAF),
in Newport News, Virginia, USA, provides 
high power electron beams of energy up to 6 GeV and 100$\%$ duty cycle to 
three experimental Halls (A, B, C) simultaneously. The CLAS spectrometer, located in Hall-B, 
is composed of several sub-detectors and two magnets. The spectrometer was 
designed to track charged particles of momentum greater than 250 MeV/c as well 
as detecting real photons. Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) used for the tracking of charged 
    particles~\cite{DCref} (in blue on the figure).
 \item Superconducting toroidal magnet (in yellow) bends the trajectories 
    of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Cherenvkov Counters (CC) used to distinguish electrons from other negative 
    particles, such as pions, at lower energies~\cite{CCref} (in pink).
 \item Scintillation Counters (SC) used to identify hadrons by measuring their 
    time of flight (TOF)~\cite{TOFref} (in red).
 \item Electromagnetic Calorimeters (EC) used to distinguish electrons from other negative 
    particles at higher energies and to detect neutral particles (neutrons and 
    photons in particular)~\cite{ECref} (in green).
\end{itemize}

@@

Measurement of the DVCS process ($eH \rightarrow e' H' \gamma$, where 
$H$ is a hadron) necessitates some adaptations in this setup.
With a 6 GeV electron beam, the 
majority of DVCS photons are produced at very forward angles where the  
acceptance of the EC is poor. To extend the range, an additional 
Inner Calorimeter (IC) was built for the E01-113 experiment in 2005. The IC is 
constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$. Each crystal is 16 cm long,
corresponding to 17 radiation lengths. The achieved energy 
resolution is around 3 to 4$\%$ for photon energies between 2 GeV and 5 GeV, with
an angular resolution between 3 and 5 mrad 
\cite{Hyon-suk}.

%M{\o}ller electrons can be
%very problematic at such small angles, so a 5 Tesla solenoid magnet has been
%added around the target as a magnetic shield against this low energy background.

For coherent nuclear processes, the RTPC was developed to extend the tracking range of CLAS to low energy nuclei and was used during a one month experimental run in 2009 
with a longitudinally polarized electron beam of energy 6.064 GeV incident 
on a gaseous $^{4}$He target.
%, with the goal of measuring asymmetries in Deep Virtual Compton 
%Scattering (DVCS)~\cite{proposal} and exclusive meson production.

Moreover, a 
5~T solenoid was placed around the RTPC-target system to shield the IC and RTPC from the large flux of low energy M{\o}ller electrons. The solenoid field serves also as the
analyzing magnet for charged particle tracking by the RTPC.

In this paper we report the design, calibration, and performance of the radial TPC, organized as follows.
Section \ref{prev_tpc} presents an overview of the history of TPCs.
Section \ref{sec_design} details the physics considerations for 
our RTPC's design and internal structure. Section \ref{sec_readout} 
describes the properties of the read-out system.  Calibration strategies are
discussed in section \ref{sec_calib}. Finally, the 
performance of the RTPC is described in section \ref{sec_perf}.

\section{TPC technology} \label{prev_tpc}
In the late 1970s, David R. Nygren presented the concept of the Time Projection 
Chambers (TPCs) as a new tool to track charged particles based on combining 
electric and magnetic fields within a sensitive gas volume. The ionized 
electrons are collected by a position-sensitive electron collection system.  
Later in 1997, F.~Sauli introduced the Gas Electron Multiplier (GEM) as an 
effective gas-gain element to improve the tracking efficiency of the TPCs 
\ref{gem_sauli}. Nowadays, the TPCs are widely used in physics experiments, 
such as high-energy physics (ALEPH, DELPHI), heavy-ion physics (STAR, ALICE) 
and rare-event searches (XENON, NEXT).

The first implementation of a radial TPC was  
performed by the BoNuS experiment \cite{BONUS} in 2005. The success of the 
BoNuS-RTPC in detecting the slow protons motivated the development of this new 
RTPC. For the channels of interest in CLAS-EG6 experiment, the recoiling 
$^{4}$He nuclei have momentum per charge around 150 MeV/c. The detection of 
these low-energy nuclei was the primary motivation for designing this detector.
